---
Title: "Cyclistic_Analysis"
Author: "Irene Li"
Date: "2021-12-15"
Output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a capstone project for my journey of Google Data Analytics Professional Certificate. 

This project is about getting insights of how bike-sharing differ between casual users and annual members for Cyclistic Company. 

The analysis workflow will follow the six-step of data analysis: Ask, Prepare, Process, Analysis, Share, and Act. 

Following are steps that illustrated how I approached my findings and what decision-making I will provide in the end. 

## Business Task

How do annual members and casual riders use Cyclistic bikes differently?


## Data Preparation

Data was made available and downloaded via Motivate International Inc. under this [liscense](https://ride.divvybikes.com/data-license-agreement). 
Data has coverage over the period from January 2020 to December 2020. The data was downloaded and unzipped as 10- .csv files, 
one presented the first quarter of 2020 and the rest - one for each month of the range. Original copies and 10 .csv files were saved as separate folders. 
I opened .csv files in Excel and saved them as Excel Workbook files, so I can perform analysis later. 
All files were saved in separate folders (.zip / .csv / .xls) properly. 

Notes: This is a public data that can be used to explore how different customer types are using Cyclistic bikes. 
But data-privacy issues prohibit users from using riders' personally identifiable information. This means no one will be 
able to connect pass purchases to credit card numbers to determine if casual riders live in Cyclistic service area or if they 
have purchased multiple single passes.

## Set up

First of all, I need to install required packages so I can perform analysis 

* tidyverse & dplyr - for data import and wrangling 
* lubridate - for data functions working with dates and time
* ggplot - for create data visualization in R

```{r libraries}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("lubridate")
install.packages("ggplot2")
```

## Call installed packages

```{r installed_packages, echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
```

## STEP 1: COLLECT DATA (UPLOAD DATA) for the previous 12 months

Data has covererage over the period from Janurary 2020 to December 2020. 
The data was downloaded and unzipped as 10- .csv files, one presented the first quarter of 2020 and the rest - one for each month of the range.

```{r original_data}
trip_q1_2020 <- read.csv("2020_Q1_trips.csv")
trip_04_2020 <- read.csv("2020_04_trips.csv")
trip_05_2020 <- read.csv("2020_05_trips.csv")
trip_06_2020 <- read.csv("2020_06_trips.csv")
trip_07_2020 <- read.csv("2020_07_trips.csv")
trip_08_2020 <- read.csv("2020_08_trips.csv")
trip_09_2020 <- read.csv("2020_09_trips.csv")
trip_10_2020 <- read.csv("2020_10_trips.csv")
trip_11_2020 <- read.csv("2020_11_trips.csv")
trip_12_2020 <- read.csv("2020_12_trips.csv")
```

## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

I checked the columns names of each file to make sure they were consistent, so I could merge them together later.

```{r columns_name}
colnames(trip_q1_2020)
colnames(trip_04_2020)
colnames(trip_05_2020)
colnames(trip_06_2020)
colnames(trip_07_2020)
colnames(trip_08_2020)
colnames(trip_09_2020)
colnames(trip_10_2020)
colnames(trip_11_2020)
colnames(trip_12_2020)
```

All columns names were checked and they were all the same. Aside of ensuring consistency of column names, I also checked the internal structure of data 
frames for any incongruities.

```{r dataframe_structures}
str(trip_q1_2020)
str(trip_04_2020)
str(trip_05_2020)
str(trip_06_2020)
str(trip_07_2020)
str(trip_08_2020)
str(trip_09_2020)
str(trip_10_2020)
str(trip_11_2020)
str(trip_12_2020)
```

From the internal structure information, type of start_station_id and end_station_id were integer, so next step was to convert type of integer 
into character, in order to merge data correctly and ensure its consistency.

```{r convert int to chr}
trip_q1_2020 <- mutate(trip_q1_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_04_2020 <- mutate(trip_04_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_05_2020 <- mutate(trip_05_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_06_2020 <- mutate(trip_06_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_07_2020 <- mutate(trip_07_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_08_2020 <- mutate(trip_08_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_09_2020 <- mutate(trip_09_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_10_2020 <- mutate(trip_10_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_11_2020 <- mutate(trip_11_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
trip_12_2020 <- mutate(trip_12_2020, start_station_id = as.character(start_station_id),end_station_id = as.character(end_station_id))
```

Now, it was ready to merge all files into one big data frame!

```{r merge_one_single_dataframe}
all_trips <- bind_rows(trip_q1_2020, trip_04_2020, trip_05_2020, trip_06_2020, trip_07_2020, trip_08_2020,
                   trip_09_2020, trip_10_2020, trip_11_2020, trip_12_2020)
```

Remove variables start_lat, start_lng, end_lat, end_lng, start_station_name, start_station_id, end_station_name, end_station_id from the new data frame. 

```{r remove_variables_not_need}
all_trips <- all_trips %>%
  select(-c(start_lat, start_lng, end_lat, end_lng, start_station_name, start_station_id, end_station_name, end_station_id))
```

# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

Inspected the new table that has been created: column names, dimensions, and summarized statistics indices of the data frame. 

```{r internal_inspection_new_dataframe}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

Since the business task focused on "How do annual members and casual riders use Cyclistic bikes differently?", 
I needed to make sure the categories, as well as their amounts, under variables rideable_type and member_casual.

```{r categories_and_amount}
ride_cat <- unique(all_trips$rideable_type)
ride_cat
user_cat <- unique(all_trips$member_casual)
user_cat
table(all_trips$rideable_type)
table(all_trips$member_casual)
```

Based on results, there were two types of users under variable member_casual: 1366575 for casual users, and 2175108 for annual members. 
There were three bike types under variable rideable_type: 70616 for classic_bike, 2966322 for docked_bike, and 504745 for electric_bike.

I wanted to add more columns into the data frame, and since there was variable started_at, I could extract some date and time information from it. 

```{r extract_information_add_columns}
all_trips$date <- as.Date(all_trips$started_at) #extract date information from the column
all_trips$day <- format(as.Date(all_trips$started_at), "%d") # format the date information for day
all_trips$month <- format(as.Date(all_trips$started_at), "%m") #format the date information for month
all_trips$year <- format(as.Date(all_trips$started_at), "%Y") #format the date information for year in 4-digit
all_trips$day_of_week <- format(as.Date(all_trips$started_at), "%A") #format the date information for day of the week using unabbreviated weekday

```

I added a column called ride_length and presented the time length in minutes. 

```{r ride_length_in_mins}
all_trips$ride_length <- difftime(all_trips$ended_at, all_trips$started_at, units = "mins")
```

After adding more columns, I checked the internal structure of the data frame again.

```{r internal_structure}
str(all_trips)
```

Data type of variable ride_length was double, and I converted it into numeric so I could run calculations on the data.

```{r convert_numeric}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

# REMOVE "BAD" DATA

For the variable ride_length, it should be greater than 0 and should not be in a negative form. 

Also, I have removed those with ride_length greater than one day (1440 minutes). 

```{r removed_bad_data}
all_trips_2 <- all_trips[!(all_trips$ride_length<=0),] # create a new data frame after removing bad data
sum(all_trips$ride_length<=0)
sum(all_trips_2$ride_length<=0)

sum(all_trips$ride_length>=1440)
all_trips_2 <- all_trips_2[!(all_trips_2$ride_length>=1440),]
sum(all_trips_2$ride_length>=1440)
```

There were 11126 rows that ride_length was equal to 0 or smaller than 0, and there were 2863 rows that ride_length was equal or greater than one day. 
All of these rows were being removed. 

At this point, there were following variables in the new data frame with those ride_length in negative form and over one day being removed:

* ride_id
* rideable_type: classic_bike, docked_bike, and electric_bike
* started_at
* ended_at
* member_casual: member, casual
* date
* month
* year
* day_of_week
* ride_length

# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

I computed some descriptive indices on ride_length: 

```{r descriptive_analysis}
mean(all_trips_2$ride_length)
median(all_trips_2$ride_length) 
min(all_trips_2$ride_length) 
max(all_trips_2$ride_length) 
summary(all_trips_2$ride_length) 
```

* Mean ride_length: 23.5301 mins
* Median ride_length: 14.1 mins
* Minimum ride_length: 0.0167 mins
* Maximum ride_length: 1439.817 mins

Besides the ride_length, I wanted to visualize the proportion of (pie chart):

* annual members and casual users over the total users population
* classic_bike, docked_bike, and electric_bike over the total bikes population

```{r proportion_member_casual}
all_trips_2 %>%
  group_by(member_casual) %>%
  summarise(number_of_users = n()) %>%
  ggplot(aes(x="", y=number_of_users, fill=member_casual)) +
  geom_col(color = "black") +
  geom_text(aes(label = number_of_users), position = position_stack(vjust = 0.5),
            show.legend = FALSE) +
  coord_polar(theta = "y")
```

Annual members occupied 61.44% and casual users were around 38.54%. 

```{r proportion_bike_types}
all_trips_2 %>%
  group_by(rideable_type) %>%
  summarise(number_of_bikes = n()) %>%
  ggplot(aes(x="", y=number_of_bikes, fill=rideable_type)) +
  geom_col(color = "black") +
  geom_text(aes(label = number_of_bikes), position = position_stack(vjust = 0.5),
            show.legend = FALSE) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Proportion of Three Different Types of Bikes")
```

Over all three bike types, there were 2.00% for classic_bike, 14.30% for electric_bike, and 83.70% for docked_bike. 

Next, I compared the annual members and casual users regarding ride_length

```{r users_ridelength}
aggregate(all_trips_2$ride_length ~ all_trips_2$member_casual, FUN = mean)
aggregate(all_trips_2$ride_length ~ all_trips_2$member_casual, FUN = median)
aggregate(all_trips_2$ride_length ~ all_trips_2$member_casual, FUN = min)
aggregate(all_trips_2$ride_length ~ all_trips_2$member_casual, FUN = max)
```

I wanted to compare average ride time by the each day for annual members and casual users, but first, looking at the data frame, 
day_of_week were out of correct of order and I would assign the actual day to the number.

```{r assign_day}
all_trips_2$day_of_week <- ordered(all_trips_2$day_of_week, levels = c("Sunday","Monday",
                                                                    "Tuesday","Wednesday",
                                                                       "Thursday","Friday",
                                                                       "Saturday"))
```

```{r average_ride_length_weekday}
aggregate(all_trips_2$ride_length ~ all_trips_2$member_casual + all_trips_2$day_of_week, FUN =mean)
```

Visualized the results about weekday and average ride time based on annual members and casual users.

```{r viz_average_ride_length_weekday}
all_trips_2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Average Duration based on Rider Type")
```

Analyzed ridership data by number_of_rides and weekday.

```{r ride_length_user_weekday}
all_trips_2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

Created data visualization on number_of_rides and weekday based on annual members and causal users.

```{r viz_total_rides_weekday}
all_trips_2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Number of Bikes based on Rider Type")
```

Analyzed and visualized the relationship between different bike types usage and user types.

```{r bike_type_user}
all_trips_2 %>%
  mutate(all_trips_2$rideable_type, label = TRUE) %>%
  group_by(rideable_type, member_casual) %>%
  arrange(member_casual, rideable_type) %>%
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  ggtitle("Different Bike Types Usage")
```

# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS

I generated a summary file, exported and saved it as .csv file for future analysis.

```{r generated_summary_report}
counts <- aggregate(all_trips_2$ride_length ~ all_trips_2$member_casual + 
                      all_trips_2$day_of_week, FUN = mean)
```

```{r exported_saved_csv}
write.csv(counts, file ="/Users/irene/Desktop/Google_Project/CSV_files/avg_ride_length.csv")
```

# FINDINGS

* Casual users occupied almost 40% of the total users population. 
* Docked bike was the most popular one (over 80%) among three bike types.
* Casual users mostly use the bike-share service on the weekends, but less often during the weekdays. 
  Number of rides from casual users on the weekdays was only half of number on the weekends. 
* Average duration from casual users were twice as much as annual members' over the whole weeks.

# CONCULSIONS / SUGGESTIONS

* Provide special membership offers (e.g lower price) for using docked bikes in order to convert huge casual docked bike users into annual members. 
* Offer "weekdays membership packages" to those who use Cyclistic bikes on weekdays. 
* Set a duration milestone and create cash-back or credit-bonus membership program to convert casual users into members because 
  they have a longer average duration than current members. 

# LIMITATION / FUTURE IMPROVEMENT

* There were 11126 rows that ride duration was equal to or less than 0 minutes, and 2863 rows that ride duration was greater than one day. 
  All of these were removed and we should figure out: 
  - whether technical errors have occurred in the system (e.g. negative ride duration records)
  - and the possibility of users rent Cyclistic over a day (e.g ride duration over 1440 minutes.)    
* There were huge number of rows that the station names were missing while the latitude and longitude information recorded. 
  I didn't remove or drop these rows, but I think the missing station name could be fixed by comparing latitude and 
  longitude information from other rows. 







